---
title: "Homework 3"
author: 'Group 8: Nguyen Nhi, Mon Daniel, Sean Paolo, Gandiboyina Ethan'
date: "2024-03-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



 step 1 the text file containing the data is called **priming.txt**. You can download the data and load them from your project directory into R as a new dataframe called priming by running the following (**make changes according to your paths and OS**):
```{r}
#we load the priming database
priming <- read.table(file = "H:/Documents/data analysis/HWA3/priming.txt",
                      stringsAsFactors = FALSE)

head(priming)
```
-step 2 Get to know the data using View(), summary(), head() and str().
``` {r}
#we apply the functions to the priming database
View(priming)
head(priming)
summary(priming)
str(priming)
```
-step 3 Look at the names of the dataframe with names(). Those aren’t very informative are they? Change the
 names to the correct values (make sure to use the naming scheme I describe in the dataset description).
```{r}
#we use the function names to check the default names of the priming database
{\color{blue} names(priming)}
#we change the column names using the colname function
{\color{blue} colnames(priming)<-c("id","sex","age","attention", "prime", "prime duration", "grandparents","walk")}
#we check the new names of the priming database
names(priming)
```
-Step 4 What was the mean participant age? Answer this in two ways. First, calculate the mean directly from
 the age column. Second, create a new vector object called age.v that contains the age data, then calculate
 the mean age from this vector. Do you get the same result?
```{r}
#we use the mean function to calculate the mean participant age
mean(priming$age)
#we repeat the calculation using the second method
age.v<-c(priming$age)
mean(age.v)
```
 I got the same result 21.996
 step 5 what was the median walking time
```{r}
#we use the median function to calculate the median walking time
median(priming$walk)
```
-step 6 How many females were there? How many males?
```{r}
#we use the table function to display the count of males and females
table(priming$sex)
```
-step 7 What percent of participants passed the attention check? (Hint: To calculate a percentage from a 0, 1
 variable, use mean())
```{r}
mean(priming$attention==1)# percentage of participants that passed the attention check
```
-step 8 Walking time is currently in seconds. Add a new column to the dataframe called walk_m that shows the
 walking time in minutes rather than seconds
```{r}
# we use the cbind function to add the walking time in minutes
priming<-cbind(priming,walk_min=(c(priming$walk)/60))
head(priming)
```
-step 9 What were the sexes of the first 10 participants?
```{r}
s10<-priming[1:10,c("id","sex")] # we select the id and sex column of the priming datasets
colnames(s10)[1]="Participant id"# we change the name of the first column
print(s10)

```
-step 10 What was the data for the 50th participant?
 Try answering these questions using one of the methods in the Examples above. The easiest method is
 Method 1. That is, first create a new dataframe object of the subsetted data, and then calculate the
 summary data from this new object.
```{r}
d50<-subset(priming,id==50) #we subset the priming dataset to by the id value
summary(d50)
```
-step 11 What was the mean walking time for the elderly prime condition?
```{r}
mean(subset(priming,prime=="elderly")$walk)# we subset the priming data set and then use the mean function on the subsetted database to show the mean walking time of elderly prime condition 
```
-step 12 What was the mean walking time for the neutral prime condition?
```{r}
mean(subset(priming,prime=="neutral")$walk) # we subset the priming data set and then use the mean function on the subsetted database to show the mean walking time for the neutral prime condition
```
-step 13 What was the mean walking time for participants less than 23 years old?
```{r}
mean(subset(priming,age<23)$walk) # we subset the priming data set and then use the mean function on the subsetted database to show the mean walking time for participants less than 23 years old
```
-step 14 What was the mean walking time for females with a close relationship with their grandparents?
```{r}
print(mean(subset(priming,sex=="f" & grandparents=="yes")$walk)) # we subset the priming data set and then use the mean function on the subsetted database to show the mean walking time for females with a close relationship with their grandparents
```
-step 15 What was the mean walking time for males over 21 years old without a close relationship with their grandparents?
```{r}
print(mean(subset(priming,sex=="m" & (grandparents=="no"| grandparents=="none") & age>21)$walk))# we subset the priming data set and then use the mean function on the subsetted database to show the mean walking time for males over 21 years old without a close relationship with their grandparents
```
 
-step 16 One of your colleagues wants the study data, but only the columns id, prime, and walk. Create a new
 dataframe called priming_simple that only contains these columns.
```{r}
priming_simple<- priming[,c(1,5,8)] #we choose the 3 desired columns of the priming dataset
head(priming_simple)
```
-step 17 Some of the data don’t make any sense. For example, some walking times are negative, some prime
 values aren’t correct, and some prime.duration values weren’t part of the original study plan. Create a
 new dataframe called priming_c (aka., priming clean) that only includes rows with valid values for each
 column – do this by looking for an few strange values in each column, and by looking at the original dataset
 description. Additionally, only include participants who passed the attention check.
```{r}
colnames(priming)[6]="prime.duration"
priming_c<-subset(priming,subset = sex %in% c("f","m") &
                   age > 18 &
                   attention %in% c(0,1) &
                   prime %in% c("elderly","neutral") &
                   prime.duration %in% c(1,5,10,30) &
                   grandparents %in% c("yes","no","none") &
                   walk > 0) # we subset the priming dataset in order to remove the rows with values outside of the desired intervals
head(priming_c)
```
-step 18 How many participants gave valid data and passed the attention check? (Hint: Use the result from your previous answer!)
```{r}
table(priming_c$attention)# we get the count of the the participants who have passed and not passed the attention check
# the number of participants that passed is 291
```
-step 19 Of those participants who gave valid data and passed the attention check, what was the mean walking
 time of those given the elderly and neutral prime (calculate these separately).
```{r}
# we subset the priming_c data set and then use the mean function on the subsetted database to show the mean walking time of those given the elderly and neutral prime
mean(subset(priming_c, attention==1 & prime=="elderly")$walk)
mean(subset(priming_c, attention==1 & prime=="neutral")$walk)
```
