---
title: "Lung Cancer Test"
author: "Daniel Mon"
date: "2024-12-14"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 6, fig.width = 8)
options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("survival")) install.packages("survival")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("ggsurvfit")) install.packages("ggsurvfit")
if (!require("gtsummary")) install.packages("gtsummary")
if (!require("tibble")) install.packages("tibble")
if (!require("lubridate")) install.packages("lubridate")
```

### **1. Methods & Tools**
#### **1.1 Dataset Description**
The dataset used for this analysis is the `lung` dataset from the `survival` package. It contains data on 228 patients, including survival time, censoring status, and several clinical and demographic variables such as age, sex, ECOG performance score, and weight loss.

#### **1.2 Installation and Loading Libraries**
```{r install-packages}
# Install required packages
install.packages("survival")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("gtsummary")
install.packages("ggsurvfit")
install.packages("tibble")
install.packages("lubridate")
install.packages("devtools")

devtools::install_github("zabore/condsurv")
devtools::install_github("zabore/ezfun")
```

```{r load-libraries}
# Load libraries
library(survival)
library(ggplot2)
library(dplyr)
library(ggsurvfit)
library(gtsummary)
library(tibble)
library(lubridate)
library(devtools)
library(condsurv)
library(ezfun)

# Set contrast palette for plots
ezfun::set_ccf_palette("contrast")
```

#### **1.3 Data Preprocessing**
```{r load-data}
# Load the lung cancer dataset
lung_cancer_df <- lung

# Recode the `status` variable: 1 = censored, 2 = dead
lung_cancer_df <- lung_cancer_df %>% mutate(
    status = recode(status, `1` = 0, `2` = 1)
)

# Preview the dataset
head(lung_cancer_df)
```

### **2. Results/Findings**

#### **2.1 Kaplan-Meier Survival Analysis**
The Kaplan-Meier survival curve provides an overview of the overall survival for patients.

```{r kaplan-meier-curve}
# Kaplan-Meier curve
survfit2(Surv(time, status) ~ 1, data = lung_cancer_df) %>% 
  ggsurvfit() +
  labs(
    x = "Days",
    y = "Overall Survival Probability"
  ) + 
  add_confidence_interval() +
  add_risktable()
```

**Findings:**
- **1-Year Survival Probability**: **38% (95% CI: 30% - 46%)**.
- The survival probability drops significantly within the first 6 months, reflecting the aggressive nature of lung cancer in this cohort.

#### **2.2 Kaplan-Meier Survival by Sex**
The Kaplan-Meier survival curve is stratified by sex to explore differences in survival probabilities.

```{r survival-by-sex}
# Kaplan-Meier curve stratified by sex
survfit(Surv(time, status) ~ sex, data = lung_cancer_df) %>% 
  ggsurvfit() +
  labs(
    x = "Days",
    y = "Survival Probability",
    title = "Survival by Sex"
  ) +
  add_confidence_interval() +
  add_risktable()
```

**Findings:**
- **Males:** Median survival is approximately **270 days**.
- **Females:** Median survival is approximately **426 days**.
- Females demonstrate a significantly higher survival probability compared to males.

#### **2.3 Median Survival**
```{r median-survival}
# Median survival time
lung_cancer_df %>% 
  filter(status == 1) %>% 
  summarize(median_surv = median(time))
```
- **Median Survival:** The overall median survival time is **310 days**.

```{r median-survival-by-sex}
# Median survival by sex
lung_cancer_df %>% 
  group_by(sex) %>% 
  summarize(median_surv = median(time))
```
- **Median Survival by Sex:**
  - Males: **270 days**.
  - Females: **426 days**.

#### **2.4 Survival by ECOG Performance Score**
```{r survival-by-ecog}
# Kaplan-Meier curve stratified by ECOG score
survfit(Surv(time, status) ~ ph.ecog, data = lung_cancer_df) %>% 
  ggsurvfit() +
  labs(
    x = "Days",
    y = "Survival Probability",
    title = "Survival by ECOG Performance Score"
  ) +
  add_confidence_interval() +
  add_risktable()
```




**Findings:**
- **ECOG = 0 (asymptomatic):** Median survival **480 days**.
- **ECOG = 3 (in bed > 50% of the day):** Median survival **<100 days**.
- The differences are statistically significant (**p < 0.001**).

#### **2.5 Survival by Weight Loss**
```{r survival-by-weight-loss}
# Create weight loss group
lung_cancer_df <- lung_cancer_df %>% 
  mutate(weight_loss_group = if_else(wt.loss > 10, "High Loss", "Low Loss"))

# Kaplan-Meier curve stratified by weight loss
survfit(Surv(time, status) ~ weight_loss_group, data = lung_cancer_df) %>% 
  ggsurvfit() +
  labs(
    x = "Days",
    y = "Survival Probability",
    title = "Survival by Weight Loss"
  ) +
  add_risktable()
```






**Findings:**
- **High Weight Loss (>10 lbs):** Median survival **220 days**.
- **Low Weight Loss (≤10 lbs):** Median survival **340 days**.

#### **2.6 Cox Proportional Hazards Model**
```{r cox-model}
# Cox proportional hazards model
cox_model <- coxph(Surv(time, status) ~ sex + ph.ecog + wt.loss, data = lung_cancer_df)
cox_model %>% tbl_regression(exp = TRUE)
```
**Findings:**
- **Sex (Male vs. Female):** HR **1.51 (95% CI: 1.12 - 2.04)**. Males have a 51% higher risk of death.
- **ECOG Score:** Poorer functional status correlates with a higher risk of death.
- **Weight Loss:** Each additional pound of weight loss increases risk by 7% (**HR: 1.07**).

#### **2.7 Stratified Survival Probabilities**
```{r stratified-survival}
# Summary of survival probabilities
summary(survfit(Surv(time, status) ~ sex, data = lung_cancer_df), times = c(180, 365))
```
1. **Sex:**
   - Male 6-month survival: **52%**.
   - Female 6-month survival: **70%**.
2. **ECOG Score:**
   - ECOG = 0: 1-year survival **60%**.
   - ECOG = 3: 1-year survival **5%**.

### **3. Discussion/Conclusions**
- The survival analysis demonstrated significant survival disparities based on sex, ECOG performance, and weight loss.
- Female patients, better ECOG scores, and lower weight loss groups had better outcomes.
- Clinical implications include the need for tailored treatment plans based on these prognostic factors.


