---
title: "R markdown example"
author: "Insert here with full name"
date: "`r format(Sys.time(), '%a %d %b %Y')`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Main Title 1

# underlined Title 2
*** 

## Header 2 

### Header 3
  
* Item 1
* Item 2
+ Item 2a
+ Item 2b


1. Item 1
2. Item 2
3. Item 3
+ Item 3a
+ Item 3b


```{r include=FALSE}
#```{r}
library(dbplyr)
library(ggplot2)
```

**Part1**
```{r}
data("cars")
str(cars)
summary(cars$dist)
summary(cars$speed)
plot(cars)
```

Note: a keyboard shortcut to produce code chunk brackets in Rmarkdown is:
  Ctrl + Alt + i

**Part2**
  
This is a *simple plot* for *italics*
```{r}
hist(cars$speed)
boxplot(cars$dist)
```

This is a **customized plot** for **BOLD**
  
```{r fig.width=15, fig.height=10}
boxplot(cars$dist)
```

There were `r nrow(cars)` cars studied.  

The mean spead of cars is `r mean(cars$speed)`.

Here is a LaTeX equation in line: $\sum_{n=1}^{10} n^2$
  
Here an equation centered: $$\sum_{n=1}^{10} n^2$$
  
> Here is a single block quote in this report...


**Chunk Options**

Chunk output can be customized with knitr options, arguments set in the {} of a chunk header. Above, we use five arguments:

- **include = FALSE** prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.
- **echo = FALSE** prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.
- **message = FALSE** prevents messages that are generated by code from appearing in the finished file.
- **warning = FALSE** prevents warnings that are generated by code from appearing in the finished.
- **fig.cap = "..."** adds a caption to graphical results.

See the R Markdown Reference Guide for a complete list of knitr chunk options.

**global Options**

To set global options that apply to every chunk in your file, call `knitr::opts_chunk$set` in a code chunk. Knitr will treat each option that you pass to `knitr::opts_chunk$set` as a global default that can be overwritten in individual chunk headers.



# .. the end ...
