Certainly! Below is an R Markdown (Rmd) file that performs linear regression manually and calculates the equation for the line of best fit. You can copy this content into an R Markdown file and run it to see the results.

```markdown
---
title: "Linear Regression: Line of Best Fit"
author: "Your Name"
date: "Today's Date"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Linear Regression: Finding the Line of Best Fit Manually

Given data:
- \(X\) values: \(0, 3, 2, 1, 5, 4, 3, 0, 4, 5\)
- \(Y\) values: \(2, 8, 7, 3, 12, 10, 9, 3, 9, 13\)

### Step-by-Step Calculation

```{r}
# Data
X <- c(0, 3, 2, 1, 5, 4, 3, 0, 4, 5)
Y <- c(2, 8, 7, 3, 12, 10, 9, 3, 9, 13)

# Number of data points
n <- length(X)

# Calculate means
X_bar <- mean(X)
Y_bar <- mean(Y)

# Calculate the components for the slope (m)
numerator <- sum((X - X_bar) * (Y - Y_bar))
denominator <- sum((X - X_bar)^2)

# Calculate the slope (m) and intercept (b)
m <- numerator / denominator
b <- Y_bar - m * X_bar

# Display the results
m
b
```

The slope (\(m\)) of the line of best fit is `r round(m, 2)`, and the intercept (\(b\)) is `r round(b, 3)`.

### Equation of the Line of Best Fit

The equation of the line of best fit is:

\[ y = `r round(m, 2)`x + `r round(b, 3)` \]

### Plotting the Data and the Line of Best Fit

```{r}
# Plot the data points
plot(X, Y, main = "Scatter Plot with Line of Best Fit", xlab = "X", ylab = "Y", pch = 19, col = "blue")

# Add the line of best fit
abline(a = b, b = m, col = "red", lwd = 2)

# Add legend
legend("topleft", legend = c("Data points", "Line of Best Fit"), col = c("blue", "red"), pch = c(19, NA), lty = c(NA, 1), lwd = c(NA, 2))
```

In this plot, the blue points represent the original data, and the red line represents the line of best fit calculated manually.
```

This R Markdown file includes the data, the manual calculation of the slope and intercept, and a plot showing the data points and the line of best fit. After copying this into an `.Rmd` file, you can knit the document to see the results.