---
title: 'Data Analysis: Z-score calculation with R'
author: 'MANOS Thanos'
date: "`r format(Sys.time(), '%a %d %b %Y')`"
output:
  pdf_document: 
        latex_engine: xelatex
  html_document: default
  word_document: default
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(message=FALSE, echo = TRUE, tidy.opts=list(width.cutoff=60)) 
```
 
Standard score or z-score is a measure of standard deviations that how much below or above the element is located from the mean value. Z-scores are usually located around -3 to 3 sigma range (based on the variance of data, it might be different). Z-scores mean value is very close to 0, and both variance and standard deviation are equal to 1.

Z-score can be calculated with the formula below:

$$z=(x- \mu)/ \sigma $$ 

where:

   - $x$ - vector (elements of $x$ vector)
   
   - $μ$ - mean value of $x$ vector
   
   - $σ$ - standard deviation of $x$ vector

The normal distribution curve can easily explain a z-score. Z-score values are located around the curve below. Zero is a mean center value. The highest and lowest values can be found in the right and left most parts of the curve.


Let's generate some sample data and get its z-scores.

```{r, echo=T, eval=T}
set.seed(123)
```

```{r, echo=T, eval=T}
x = sample(1:50, 100, replace=T)
hist(x)
```

Getting z-scores with a formula.

```{r, echo=T, eval=T}
m = mean(x)
s = sd(x)

zs = (x - m)/s
```

```{r, echo=T, eval=T}
summary(x)
```

```{r, echo=T, eval=T}
summary(zs)
```

As summary shows, the x vector was centered into 0 mean value. In 'zs', the value of x vector's 1 is equal to -1.71, and 50 is equal to 1.84 sigma value.

In R, we can use **scale()** command to get z-scores.

```{r, echo=T, eval=T}
scale(x)
```

We need the **first part** of a scale function result.

```{r}
sc_zs = scale(x)[,1]
summary(sc_zs)
```

A summary shows that the result is the same as the one that was taken with a formula.

The scale function is often used to clean up data to remove the mean value from the series data.